---
title: "SCEPTRE example"
output: BiocStyle::html_document
date: "22/07/2024"
author: "Jana Braunger"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
BiocStyle::html_document 
data_dir = '../../example_data/'
```

# Using crispat assignment in SCEPTRE
Since SCEPTRE is tailored for the analysis of single-cell CRISPR screens, we recommend using SCEPTRE for comparing the assignment effects on downstream analysis. In this tutorial, we show how an assignment obtained by crispat can be used in the SCEPTRE pipeline. For more details on how to create a SCEPTRE object, as well as the full analysis pipeline and settings we refer to the [SCEPTRE handbook](https://timothy-barry.github.io/sceptre-book/).
 
```{r load_libraries, message = FALSE, warning = FALSE}
library(tidyverse)
library(sceptre)
library(Seurat)
library(Matrix)
```


## Load SCEPTRE object
For demonstration purposes, we start this tutorial with an already created SCEPTRE object with set analysis parameters. For details on how to get your own data into a SCEPTRE object, check out the [SCEPTRE handbook](https://timothy-barry.github.io/sceptre-book/). The analysis scripts used for the analysis shown in our paper can be found [here](https://github.com/velten-group/crispat_analysis).
```{r}
sceptre_object <- readRDS(paste0(data_dir, 'sceptre_output/sceptre_object.rds'))
```


## Replace guide assignment
Instead of running the guide assignment step with SCEPTRE, we input an assignment obtained by crispat e.g. the one with a Gaussian mixture model here: 
```{r}
# the following helper function is a slightly modified version of the original SCETPRE function 
# (modified such that it doesn't have to use a C++ function but computes the grnas_per_cell based on the perturbation_df)
process_initial_assignment_list <- function(initial_assignment_list, grna_target_data_frame, n_cells, low_moi, maximum_assignment, perturbation_df) {
  # 1. compute the vector of grnas per cell
  cell_indices <- data.frame('cell_index' = seq(n_cells))
  perturbations <- perturbation_df %>% 
    subset(select = c('cell_index', 'gRNA')) %>% 
    distinct() %>% 
    group_by(cell_index) %>% 
    summarise('n_perturbations' = n()) %>%
    right_join(cell_indices) %>% 
    mutate(n_perturbations = ifelse(is.na(n_perturbations), 0, n_perturbations)) %>%
    arrange(cell_index)
  grnas_per_cell <- perturbations$n_perturbations
  
  # 2. determine the cells that contain multiple grnas (if in low MOI and not using maximum_assignment)
  if (low_moi && !maximum_assignment) {
    cells_w_multiple_grnas <- which(grnas_per_cell >= 2L)
  } else {
    cells_w_multiple_grnas <- integer()
  }
  # 3. pool together targeting gRNAs via the or operation
  targeting_grna_group_data_table <- grna_target_data_frame |>
    dplyr::filter(grna_group != "non-targeting") |> data.table::as.data.table()
  targeting_grna_groups <- targeting_grna_group_data_table$grna_group |> unique()
  grna_group_idxs <- lapply(targeting_grna_groups, function(targeting_grna_group) {
    curr_grna_ids <- targeting_grna_group_data_table[
      targeting_grna_group_data_table$grna_group == targeting_grna_group,]$grna_id
    initial_assignment_list[curr_grna_ids] |> unlist() |> unique()
  }) |> stats::setNames(targeting_grna_groups)
  # 4. obtain the individual non-targeting grna idxs
  nontargeting_grna_ids <- grna_target_data_frame |>
    dplyr::filter(grna_group == "non-targeting") |> dplyr::pull(grna_id)
  indiv_nt_grna_idxs <- initial_assignment_list[nontargeting_grna_ids]
  # 5. construct the grna_group_idxs list
  grna_assignments_raw <- list(grna_group_idxs = grna_group_idxs,
                               indiv_nt_grna_idxs = indiv_nt_grna_idxs)
  # 6. compute the number of targeting grna groups per cell
  out <- list(grna_assignments_raw = grna_assignments_raw,
              grnas_per_cell = grnas_per_cell,
              cells_w_multiple_grnas = cells_w_multiple_grnas)
  return(out)
}

```


```{r}
# Read in crispat assignment and filter for cells with single gRNA (relevant for low MOI only)
assignment <- read.csv(paste0(data_dir, 'guide_assignments/gauss/assignments.csv')) %>% 
  group_by(cell) %>% 
  mutate(n_grnas = n()) %>% 
  filter(n_grnas == 1)

# Since SCEPTRE numbers the cells instead of using their names, we have to add indices 
# to the assignments (numbering according to the order in the gRNA count matrix)
grna_counts <- read.csv(paste0(data_dir, 'Schraivogel/grna_counts.csv')) 
cells <- gsub("\\.", "-", colnames(grna_counts)[-1])
cell_indices <- data.frame(cell_index = seq(1, length(cells)),
                           cell = cells)
assignment <- assignment %>%
  inner_join(cell_indices)

# convert df to a list
assignment_list <- split(assignment$cell_index, assignment$gRNA)
# add gRNAs without any assigned cells to the list
null_grnas <- filter(grna_counts, ! gRNA %in% assignment$gRNA)$gRNA
assignment_list <- c(assignment_list, 
               setNames(lapply(rep(list(integer(0)), length(null_grnas)), function(x) x), null_grnas))

# get processed assignment outputs
processed_assignment_out <- process_initial_assignment_list(initial_assignment_list = assignment_list,
                                                            grna_target_data_frame = sceptre_object@grna_target_data_frame,
                                                            n_cells = ncol(sceptre_object@grna_matrix[[1]]), 
                                                            low_moi = sceptre_object@low_moi,
                                                            maximum_assignment = FALSE, 
                                                            perturbation_df = assignment)

# write guide assignment results to the sceptre object
sceptre_object@grna_assignments_raw <- processed_assignment_out$grna_assignments_raw
sceptre_object@grnas_per_cell <- as.integer(processed_assignment_out$grnas_per_cell)
sceptre_object@cells_w_zero_or_twoplus_grnas <- processed_assignment_out$cells_w_multiple_grnas
sceptre_object@initial_grna_assignment_list <- assignment_list 
sceptre_object@last_function_called <- "assign_grnas"
sceptre_object@functs_called["assign_grnas"] <- TRUE
sceptre_object@grna_assignment_method = "own_assignment"
sceptre_object@cells_in_use <- sort(unique(assignment$cell_index))
```


## SCEPTRE analysis
After adding the crispat assignment to your SCEPTRE object, you can follow the remaining SCEPTRE pipeline including quality control, calibration, power check and discovery analysis setting various input parameters as needed for your data (see the [SCEPTRE handbook](https://timothy-barry.github.io/sceptre-book/)). 
